Correct and Secure Web Programming using Dependent Types and Embedded Domain-Specific Languages

### 2013-07-29 paper/main.tex
Touchups on paper

\begin{abstract}
Dependently-typed languages allow types to be used to impose arbitrarily-specific constraints on programs, in turn allowing for easier reasoning about the operation of such programs. Stronger specifications do however bring with them the disadvantage that it becomes increasingly difficult to write programs that are accepted by the type checker, and additional proofs may have to be specified by a user.

Research into embedded domain-specific languages attempt to address this problem by introducing a layer of abstraction over more specific underlying types, allowing for domain-specific code to be written in high-level languages which use dependent types to enforce certain invariants without additional proof obligations.

In this paper, we apply this technique to web programming, showing that embedded domain-specific languages can be constructed to enforce resource usage protocols associated with common web operations such as CGI, database access and session handling. Additionally, we show that this technique may be employed to allow the creation and handling of web forms which retain type information, reducing the scope for programmer error and attacks which exploit the loss of type information, such as SQL injection.
\end{abstract}

@@

\section{Introduction}
Web applications, whilst ubiquitous, are also prone to incorrect construction and security exploits. The majority of web applications are written in dynamically-checked scripting languages \cite{}, which allow rapid development but at the same time do not give the same static guarantees about runtime behaviour afforded by programs with more expressive, static type systems, instead relying on extensive unit testing to ensure correctness and security.

@@

There is, however, no way to statically guarantee that this protocol is adhered to. A developer may, for example, accidentally close a statement whilst still retrieving rows, which would cause a runtime error. Similarly, a programmer may omit closing the statement or connection, and can lead to substantial problems such as handle leaks in longer-running server applications.

@@

\section{An overview of the \texttt{Effects} framework}
\label{effects}
\texttt{Effects} is an EDSL which handles side-effects such as database access as \textit{algebraic effects}, allowing for greater code modularity. This section is intended to serve as a brief introduction to the framework; for more information, see \cite{brady:effects}.

@@

By modelling CGI as an algebraic effect, we may enforce a resource usage protocol which, even though the program may be running in an IO execution context, prevents arbitrary IO from being performed and therefore ensures that the headers are firstly written correctly. In order to accomplish this, we define an effect, \texttt{Cgi}, and an associated resource, \texttt{InitialisedCGI}, which is parameterised over the current state, \texttt{CGIStep}. This resource describes the current state, alongside a \texttt{CGIInfo} record which contains information from the request. We represent an uninitialised CGI effect as the unit type, ().
{\small
\begin{verbatim}
data CGIStep = Initialised 
             | TaskRunning 
             | TaskCompleted 
             | HeadersWritten 
             | ContentWritten

@@

{\small
\begin{verbatim}
data Cgi : Effect where
    Init : Cgi () (InitialisedCGI Initialised) ()
    StartRun : Cgi (InitialisedCGI Initialised) 
                   (InitialisedCGI TaskRunning) ()
    FinishRun : Cgi (InitialisedCGI TaskRunning) 
                    (InitialisedCGI TaskCompleted) ()
    WriteHeaders : Cgi (InitialisedCGI TaskCompleted) 
                       (InitialisedCGI HeadersWritten) ()
    WriteContent : Cgi (InitialisedCGI HeadersWritten) 
                       (InitialisedCGI ContentWritten) ()
    OutputData : String -> 
                 Cgi (InitialisedCGI TaskRunning) 
                     (InitialisedCGI TaskRunning) ()
    RunAction : Env IO (CGI (InitialisedCGI TaskRunning) 
                               :: effs) -> 
                       CGIProg effs a -> 
                       Cgi (InitialisedCGI TaskRunning) 
                           (InitialisedCGI TaskRunning) a
    ...

\end{verbatim}
}
User code executes in the \texttt{TaskRunning} state. Several operations, such as querying the POST and GET variables, are available in this state, alongside functions to output data to the web page and append data to the response headers. It is important to note, however, that at this stage nothing is written to the page, with the \texttt{output} and \texttt{addHeader} functions instead modifying the CGIInfo record so that it can be printed at the end of the program's execution, in accordance with the resource usage protocol.

@@

By incorporating pointers to open connections and prepared statements into the resource associated with the effect, we introduce a further layer of abstraction, hiding implementation details from the developer and encouraging cleaner and less error-prone code.

\subsubsection{Example Code}
Programs making use of the DSL should look familiar to developers even without a background in functional programming. To demonstrate the functionality of the library, we present a simple example wherein a database is queried in order to enumerate the contents of a simple table. Due to the fact that the \texttt{Effects} library overloads the bind operator, we may make use of \texttt{do}-notation, facilitating the usage of an imperative style.

Firstly, we define a function of type:
{\small
\begin{verbatim}
Eff IO [SQLITE ()] (Either String (List (String, Int)))
\end{verbatim}
}
This means that the program will be run in the IO execution context, and must start and end with no active resources. The return type indicates that either a list of \text{(String, Int)} pairs, representing names and ages in the database, or an error will be returned. 
{\small
\begin{verbatim}
testSelect : Eff IO [SQLITE ()] 
             (Either String (List (String, Int)))
testSelect = do
  open_db <- openDB "people.db"
  if open_db then do
    let sql = "SELECT * FROM `test`;"
    sql_prep_res <- prepareStatement sql
    if sql_prep_res then do 
      startBind
      finishBind
      beginExecution
      results <- collectResults
      finaliseStatement
      closeDB
      return $ Right results
    else do err <- stmtFail
            return $ Left err
  else do err <- connFail
          return $ Left err $
\end{verbatim}
}
The program firstly attempts to open a connection to the ``people.db'' database. At this point, since the \texttt{OpenDB} operation has been invoked, the program transitions to the \texttt{ConnectionOpened} state. The \texttt{openDB} function returns a Boolean value indicating whether or not the operation is successful. If not, then the \texttt{connFail} function is called to generate an appropriate error and dispose of the resources.

@@

In order to decrease unnecessary boilerplate code in user applications, we provide functions which abstract out unnecessary parts of this process. In order to do this, we define the algebraic data type \texttt{DBVal}, which is a tagged union over simple primitive types:
{\small
\begin{verbatim}
data DBVal = DBInt Int
           | DBText String
           | DBFloat Float
           | DBNull
\end{verbatim}
}
We also make use of the \texttt{ResultSet} type, which is a list of returned database rows.
\begin{verbatim}
ResultSet : Type
ResultSet = List (List DBVal)
\end{verbatim}
Using these, we may implement a function, \texttt{ExecuteSelect}, which, given a query, a list of variables to bind and their associated indices within the query, and a function which is used to extract information out of each returned database row, returns either a \texttt{ResultSet} or an error.
{\small
\begin{verbatim}
executeSelect : String ->
                String -> 
                List (Int, DBVal) -> 
                (Eff m [SQLITE 
                  (SQLiteRes PreparedStatementExecuting)
                   ] 
                 (List DBVal)
                ) -> 
                Eff m [SQLITE ()] (Either String ResultSet)
\end{verbatim}
}
% =================================================

@@

Much akin to the SQLite effect, we encapsulate failure by reflecting it in the resource associated with the effect. 
{\small
\begin{verbatim}
data SessionStep = SessionUninitialised
                 | SessionInitialised

@@

{\small
\begin{verbatim}
SELECT `name`, `age` FROM `staff` LIMIT ' + page + ', 5
\end{verbatim}
}
The \texttt{page} variable is assumed to be an integer, but may instead be modified by an attacker to include a malicious string which would alter the semantics of the query, allowing an attacker to execute a blind SQL injection attack. % Might be a good idea to cite an SQL injection paper which uses LIMIT clauses here

@@

    Submit : (mkHandlerFn ((reverse G), E)) ->
             String -> 
             Form (FormRes G E) (FormRes [] []) String
\end{verbatim}
}
The implementation of the form EDSL also contains other constructs for additional form elements such as radio buttons and check boxes, but these are omitted here for brevity.

@@

Before associating a handler function with the form, we must firstly specify the effects available to the handler. This is done through the use of the \texttt{useEffects}, which updates the list of effects in the type of the form resource. By doing this, we may subsequently use the effects in calculations at the type level, in particular when calculating the type of the handler function for the form. 
{\small
\begin{verbatim}
  useEffects : (effs : List WebEffect) ->
               EffM m [FORM (FormRes G E)] 
                      [FORM (FormRes G effs)] ()
  useEffects effs = (UseEffects effs)
\end{verbatim}}
Whilst it is not possible to serialise arbitrary effects due to the associated difficulties with serialising initial resource environments, we allow for three effects to be serialised: \texttt{CGI}, \texttt{SQLITE} and \texttt{SESSION}. This is, however, not an inherent limitation as the \texttt{Effects} library allows for the introduction of effects within an effectful computation.
%
We may specify a handler function of type \texttt{FormHandler}:
{\small
\begin{verbatim}
  FormHandler : List EFFECT -> Type
  FormHandler effs = Eff IO effs ()
\end{verbatim}
}
In order to associate a handler with a form, we may call the \texttt{addSubmit} function:
{\small
\begin{verbatim}
  addSubmit : (f :  mkHandlerFn ((reverse G), E)) ->
              (fns : HandlerList) ->
              {default tactics { applyTactic 
                findFn 100; solve; }
               prf : FnElem f fns} ->
              EffM m [FORM (FormRes G E)]
                     [FORM (FormRes [] [])] 
                     String
  addSubmit f handlers {prf} = (Submit f name)
    where name : String
          name = getString' f handlers prf          
\end{verbatim}
}
Let us look at each aspect of this function in turn. Firstly, the \texttt{mkHandlerFn} function calculates the required type of the handler function from the list of types associated with the form elements, and the effects we specified with the \texttt{useEffects} function. Note that since we prepend types to the list of \texttt{FormTy}s as opposed to appending them, we must reverse the list.
{\small
\begin{verbatim}
MkHandlerFnTy : Type
MkHandlerFnTy = (List FormTy, List WebEffect)

@@

To store a reference to a handler function, we use the \texttt{HandlerFn} type:
\begin{verbatim}
public
HandlerFn : Type
HandlerFn = (ft ** (mkHandlerFn ft, String))
\end{verbatim}
%
The \texttt{**} notation denotes a dependent pair, wherein one argument, in this case the concrete handler function, depends on another, namely \texttt{MkHandlerFnTy} data used to construct the type of the handler function. We also store a unique string identifier, which is used to serialise a reference to the handler function. We then store each \texttt{HandlerFn} in a \texttt{HandlerList}.

\begin{verbatim}
public
HandlerList : Type
HandlerList = List HandlerFn
\end{verbatim}
To enforce the requirement that a supplied handler function must reside in the list of available handlers, and therefore allow us to retrieve the name with which to serialise the handler, we require a \textit{list membership proof},  \texttt{FnElem f fns}, which statically guarantees that a given item resides in a list.
{\small
\begin{verbatim}
  data FnElem : mkHandlerFn ((reverse G), E) -> 
                HandlerList -> Type where

@@

\subsection{Form Handling}
\label{formhandling}
Once the form has been submitted, a web application may handle the submitted data by invoking the \texttt{HandleForm} CGI operation. This will firstly check for the existence of the hidden \texttt{handler} field, which contains the previously serialised metadata about the form handling function, before deserialising the data into a \texttt{MkHandlerFnTy}.

With this data, we may then look up the function in the list of registered handlers by using the serialised unique handler identifier. In order to apply the handler function to the data submitted in the form, we must firstly prove to the type-checker that the deserialised \texttt{MkHandlerFnTy} is the same as the one retrieved from the list of registered handlers. We do this by making use of the \texttt{decEq} function, which determines whether two types are equal, returning a proof of equality if so, and a proof of inequality if not.

@@

\section{Extended Example: Message Board}
\label{messageboard}
Using these components, we may now consider a larger example of a message board, allowing users to register, log in, view and create threads, and list and create new posts in threads.

@@

{\small
\begin{verbatim}
main : IO ()
main = do (runCGI [initCGIState, 
                  InvalidSession, ()] 
           handleRequest)
          pure ()
\end{verbatim}
}
The \texttt{handleRequest} function firstly checks to see whether the CGI request must handle previously-submitted form data, via the \texttt{isHandlerSet} function. If so, then the form handling routine is called, which handles the form data as described in Section ~\ref{formhandling}. If not, then the \texttt{handleNonFormRequest} function is called, which inspects the GET variables in order to display the correct page.
{\small
\begin{verbatim}
handleRequest : CGIProg [SESSION 
  (SessionRes SessionUninitialised), 
  SQLITE ()] ()
handleRequest = do 
  handler_set <- isHandlerSet
  if handler_set then do
    lift' (handleForm handlers)
    Effects.pure ()
  else do
    action <- lift' (queryGetVar "action")
    thread_id <- lift' (queryGetVar "thread_id")
    handleNonFormRequest action (map strToInt thread_id)
\end{verbatim}
}
\subsection{Thread Creation}
We specify four forms: one to handle registration, one to handle logging in, one to handle the creation of new threads, and one to handle the creation of new posts. One such example would be the form used to create a new thread, which contains elements for the title of the new thread and the content of the first post of the new thread.

@@

There are firstly many other applications which make use of specific resource usage protocols, for example popular libraries such as \textit{libgcrypt} \cite{libgcrypt}. Applying a similar approach would allow for sensitive programs requiring cryptographic routines to be written using a language with full dependent types, in turn adding an extra layer of confidence to their security.

>>>>

\begin{abstract}
Dependently-typed languages allow types to be used to impose arbitrarily-specific constraints on programs, in turn allowing easier reasoning about the operation of such programs. Stronger specifications do however bring with them the disadvantage that it becomes increasingly difficult to write programs that are accepted by the type checker and additional proofs may have to be specified by a user.

Research into embedded domain-specific languages attempt to address this problem by introducing a layer of abstraction over more specific underlying types, allowing domain-specific code to be written in high-level languages which use dependent types to enforce certain invariants without additional proof obligations.

In this paper, we apply this technique to web programming, and show that embedded domain-specific languages can be constructed to enforce resource usage protocols associated with common web operations such as CGI, database access and session handling. Additionally, we show that this technique may be employed to allow the creation and handling of web forms which retain type information, reducing the scope for programmer error and attacks such as SQL injection, which exploit the loss of type information.
\end{abstract}

@@

\section{Introduction}
Web applications, whilst ubiquitous, are also prone to incorrect construction and security exploits. The majority of web applications are written in dynamically-checked scripting languages \cite{}, which allow rapid development. A significant drawback, however, is that such languages do not provide  the same static guarantees about runtime behaviour afforded by programs with more expressive, static type systems, instead relying on extensive unit testing to ensure correctness and security.

@@

There is, however, no way to statically guarantee that this protocol is adhered to. A developer may, for example, accidentally close a statement whilst still retrieving rows, which would cause a runtime error. Similarly, a programmer may omit closing the statement or connection, which can lead to substantial problems such as handle leaks in longer-running server applications.

@@

\section{An overview of the \texttt{Effects} framework}
\label{effects}
\texttt{Effects} is an EDSL which handles side-effects such as database access as \textit{algebraic effects}, allowing greater code modularity. This section is intended to serve as a brief introduction to the framework; for more information, see \cite{brady:effects}.

@@

By modelling CGI as an algebraic effect, we may enforce a resource usage protocol which, even though the program may be running in an IO execution context, prevents arbitrary IO from being performed and therefore ensures that the headers are written correctly. In order to accomplish this, we define an effect, \texttt{Cgi}, and an associated resource, \texttt{InitialisedCGI}, which is parameterised over the current state, \texttt{CGIStep}. This resource describes the current state, alongside a \texttt{CGIInfo} record which contains information from the request. We represent an uninitialised CGI effect as the unit type, ().
{\small
\begin{verbatim}
data CGIStep = Initialised 
             | TaskRunning 
             | TaskCompleted 
             | HeadersWritten 
             | ContentWritten

@@

{\small
\begin{verbatim}
data Cgi : Effect where
    Init : Cgi () (InitialisedCGI Initialised) ()
    StartRun : Cgi (InitialisedCGI Initialised) 
                   (InitialisedCGI TaskRunning) ()
    FinishRun : Cgi (InitialisedCGI TaskRunning) 
                    (InitialisedCGI TaskCompleted) ()
    WriteHeaders : Cgi (InitialisedCGI TaskCompleted) 
                       (InitialisedCGI HeadersWritten) ()
    WriteContent : Cgi (InitialisedCGI HeadersWritten) 
                       (InitialisedCGI ContentWritten) ()
    OutputData : String -> 
                 Cgi (InitialisedCGI TaskRunning) 
                     (InitialisedCGI TaskRunning) ()
    RunAction : Env IO (CGI (InitialisedCGI TaskRunning) 
                               :: effs) -> 
                       CGIProg effs a -> 
                       Cgi (InitialisedCGI TaskRunning) 
                           (InitialisedCGI TaskRunning) a
\end{verbatim}
}
Additional operations, including those to query POST and GET variables, are omitted in the interest of brevity.

User code executes in the \texttt{TaskRunning} state. Several operations, such as querying the POST and GET variables, are available in this state, alongside functions to output data to the web page and append data to the response headers. It is important to note, that at this stage nothing is written to the page, with the \texttt{output} and \texttt{addHeader} functions instead modifying the CGIInfo record. This data may then be printed at the end of the program's execution, in accordance with the resource usage protocol.

@@

By incorporating pointers to open connections and prepared statements into the resource associated with the effect, we introduce a further layer of abstraction, in order to hide implementation details from the developer and encourage cleaner, less error-prone code.

\subsubsection{Example Code}
Programs making use of the DSL should look familiar to developers even without a background in functional programming. To demonstrate the functionality of the library, we return to the previous example of selecting the names and addresses of all staff working in a given department. Due to the fact that the \texttt{Effects} library overloads the bind operator, we may make use of \texttt{do}-notation, facilitating the usage of an imperative style.

We define a function of type:
{\small
\begin{verbatim}
String ->
Eff IO [SQLITE ()] (Either String (List (String, String)))
\end{verbatim}
}
This means that the program will be run in the IO execution context, and must start and end with no active resources. The return type indicates that either a list of \text{(String, String)} pairs, representing names and addresses in the database, or an error will be returned. 
{\small
\begin{verbatim}
testSelect : String -> Eff IO [SQLITE ()] 
             (Either String (List (String, Int)))
testSelect dept = do
  open_db <- openDB "people.db"
  if open_db then do
    let sql = "SELECT name, address FROM `staff` 
                    WHERE dept = ?;"
    sql_prep_res <- prepareStatement sql
    if sql_prep_res then do 
      startBind
      bindText 1 dept
      finishBind
      beginExecution
      results <- collectResults
      finaliseStatement
      closeDB
      return $ Right results
    else do err <- stmtFail
            return $ Left err
  else do err <- connFail
          return $ Left err $
\end{verbatim}
}
The program initially attempts to open a connection to the \texttt{people.db} database. At this point, since the \texttt{OpenDB} operation has been invoked, the program transitions to the \texttt{ConnectionOpened} state. The \texttt{openDB} function returns a Boolean value indicating whether or not the operation is successful. If not, then the \texttt{connFail} function is called to generate an appropriate error and dispose of the resources, as shown in ~\ref{fig:sqlitestates}.

@@

In order to decrease unnecessary boilerplate code in user applications, we provide functions which abstract out unnecessary parts of this process. In order to do this, we define the algebraic data type \texttt{DBVal}, which is a tagged union over simple primitive types:
{\small
\begin{verbatim}
data DBVal = DBInt Int
           | DBText String
           | DBFloat Float
           | DBNull
\end{verbatim}
}
We also make use of the \texttt{ResultSet} type, which is a list of returned database rows.
{\small
\begin{verbatim}
  ResultSet : Type
  ResultSet = List (List DBVal)
\end{verbatim}
}
Using these, we may implement a function, \texttt{ExecuteSelect}, which, given a query, a list of variables to bind and their associated indices within the query, and a function which is used to extract information out of each returned database row, returns either a \texttt{ResultSet} or an error.
{\small
\begin{verbatim}
executeSelect : String ->
                String -> 
                List (Int, DBVal) -> 
                (Eff m [SQLITE 
                  (SQLiteRes PreparedStatementExecuting)
                   ] 
                 (List DBVal)
                ) -> 
                Eff m [SQLITE ()] (Either String ResultSet)
\end{verbatim}
}
% =================================================

@@

Much like the SQLite effect, we encapsulate failure by reflecting it in the resource associated with the effect. 
{\small
\begin{verbatim}
data SessionStep = SessionUninitialised
                 | SessionInitialised

@@

{\small
\begin{verbatim}
SELECT `name`, `address` FROM `staff` LIMIT ' + page + ', 5
\end{verbatim}
}
The \texttt{page} variable is assumed to be an integer, but may instead be modified by an attacker to include a malicious string which would alter the semantics of the query, allowing an attacker to execute a blind SQL injection attack. % Might be a good idea to cite an SQL injection paper which uses LIMIT clauses here

@@

    Submit : (mkHandlerFn ((reverse G), E)) ->
             String -> 
             Form (FormRes G E) (FormRes [] []) String
\end{verbatim}
}
The implementation of the form EDSL also contains other constructs for additional form elements such as radio buttons and check boxes, but are omitted in the interest of brevity.

@@

Before associating a handler function with the form, we must specify the effects available to the handler. This is done through the use of the \texttt{useEffects}, which updates the list of effects in the type of the form resource. By doing this, we may subsequently use the effects in calculations at the type level, in particular when calculating the type of the handler function for the form. 
{\small
\begin{verbatim}
  useEffects : (effs : List WebEffect) ->
               EffM m [FORM (FormRes G E)] 
                      [FORM (FormRes G effs)] ()
  useEffects effs = (UseEffects effs)
\end{verbatim}}
Whilst it is not possible to serialise arbitrary effects due to the associated difficulties with serialising initial resource environments, we allow for three effects to be serialised: \texttt{CGI}, \texttt{SQLITE} and \texttt{SESSION}. This is, however, not an inherent limitation as the \texttt{Effects} library allows for the introduction of effects within an effectful computation.
%
We may specify a handler function of type \texttt{FormHandler}:
{\small
\begin{verbatim}
  FormHandler : List EFFECT -> Type
  FormHandler effs = Eff IO effs ()
\end{verbatim}
}
In order to associate a handler with a form, we may call the \texttt{addSubmit} function:
{\small
\begin{verbatim}
  addSubmit : (f :  mkHandlerFn ((reverse G), E)) ->
              (fns : HandlerList) ->
              {default tactics { applyTactic 
                findFn 100; solve; }
               prf : FnElem f fns} ->
              EffM m [FORM (FormRes G E)]
                     [FORM (FormRes [] [])] 
                     String
  addSubmit f handlers {prf} = (Submit f name)
    where name : String
          name = getString' f handlers prf          
\end{verbatim}
}
Let us look at each aspect of this function in turn. Firstly, the \texttt{mkHandlerFn} function calculates the required type of the handler function from the list of types associated with the form elements, and the effects we specified with the \texttt{useEffects} function. Note that since we prepend types to the list of \texttt{FormTy}s as opposed to appending them, we must reverse the list.
{\small
\begin{verbatim}
MkHandlerFnTy : Type
MkHandlerFnTy = (List FormTy, List WebEffect)

@@

To store a reference to a handler function, we use the \texttt{HandlerFn} type:
{\small
\begin{verbatim}
HandlerFn : Type
HandlerFn = (ft ** (mkHandlerFn ft, String))
\end{verbatim}
}
%
The \texttt{**} notation denotes a dependent pair, wherein one argument, in this case the concrete handler function, depends on another, namely \texttt{MkHandlerFnTy} data used to construct the type of the handler function. We also store a unique string identifier, which is used to serialise a reference to the handler function. We then store each \texttt{HandlerFn} in a \texttt{HandlerList}.

\begin{verbatim}
HandlerList : Type
HandlerList = List HandlerFn
\end{verbatim}
To enforce the requirement that a supplied handler function must reside in the list of available handlers, and therefore allow us to retrieve the name with which to serialise the handler, we require a \textit{list membership proof},  \texttt{FnElem f fns}, which statically guarantees that a given item resides in a list.
{\small
\begin{verbatim}
  data FnElem : mkHandlerFn ((reverse G), E) -> 
                HandlerList -> Type where

@@

\subsection{Form Handling}
\label{formhandling}
Once the form has been submitted, a web application may handle the submitted data by invoking the \texttt{HandleForm} CGI operation. This will check for the existence of the hidden \texttt{handler} field, which contains the previously serialised metadata about the form handling function, before deserialising the data into a \texttt{MkHandlerFnTy}.

With this data, we may then look up the function in the list of registered handlers by using the serialised unique handler identifier. In order to apply the handler function to the data submitted in the form, we must first prove to the type-checker that the deserialised \texttt{MkHandlerFnTy} is the same as the one retrieved from the list of registered handlers. We do this by making use of the \texttt{decEq} function, which determines whether two types are equal, returning a proof of equality if so, and a proof of inequality if not.

@@

\section{Extended Example: Message Board}
\label{messageboard}
Using these components, we may now consider a larger example of a message board which allowsusers to register, log in, view and create threads, and list and create new posts in threads.

@@

{\small
\begin{verbatim}
main : IO ()
main = do (runCGI [initCGIState, 
                  InvalidSession, ()] 
           handleRequest)
          pure ()
\end{verbatim}
}
The \texttt{handleRequest} function checks to see whether the CGI request must handle previously-submitted form data, via the \texttt{isHandlerSet} function. If so, then the form handling routine is called, which handles the form data as described in Section ~\ref{formhandling}. If not, then the \texttt{handleNonFormRequest} function is called, which inspects the GET variables in order to display the correct page.
{\small
\begin{verbatim}
handleRequest : CGIProg [SESSION 
  (SessionRes SessionUninitialised), 
  SQLITE ()] ()
handleRequest = do 
  handler_set <- isHandlerSet
  if handler_set then do
    lift' (handleForm handlers)
    Effects.pure ()
  else do
    action <- lift' (queryGetVar "action")
    thread_id <- lift' (queryGetVar "thread_id")
    handleNonFormRequest action (map strToInt thread_id)
\end{verbatim}
}
\subsection{Thread Creation}
We specify four forms: one to handle registration, one to handle logging in, one to handle the creation of new threads, and one to handle the creation of new posts. One such example would be the form used to create a new thread, which contains elements for the title of the new thread and the content of the first post of the new thread.

@@

There are many other applications which make use of specific resource usage protocols, for example popular libraries such as \textit{libgcrypt} \cite{libgcrypt}. Applying a similar approach would allow for sensitive programs requiring cryptographic routines to be written using a language with full dependent types, in turn adding an extra layer of confidence to their security.
